A geographic routing approach for IPv6 in large-scale low-power and lossy networks

### 2015-04-06 2014_ELSEVIER_CEE_SI/paper_cee_special_issue.tex
ajustes finais

The use of LED lamps in street lighting started to be reported in the last few years, and it is being considered a very good solution to improve the energy efficiency of the system and also to reduce the maintenance cost, because of their long lifetime, high luminous efficiency and high color rendering index (CRI) \citep{CRI_2010}. Also, the addition of communication capability to the SLS decreases the energy consumption and the maintenance cost, because it allows for monitoring the status information of each lamp (state, current, voltage and power measurements, etc.) and to have better control of the system (e.g.: lamp dimming). Then, a smart SLS, based on LED lamps and a communication network, is an interesting application in a smart city context, not only by its clearly advantages but also because it can be one of the foundations for a urban ``IoT'' due to its pervasiveness across the city \citep{Smart_City_Smart_Light_2013}. Therefore, it can also be an enabler for the aforementioned applications in the smart city.

However, a street lighting system features a large number of independent devices, with geographic distribution depending on the city streets. Then, adding communication capabilities to these devices requires a complex network topology, as well as interoperability, scalability, security, robustness, low cost, ease to use and maintenance. A few contributions in the literature suggest some technologies that can be used to control a street lighting system. Some use the power lines for data transmission (PLC) \citep{PLC_Street_Lighting_2006}, while others use wireless communication, such as cellular networks \citep{Cellular_Street_Lighting_2010_1}, wireless sensor networks \citep{WSN_Street_Lighting_2010_1} and both combined \citep{WSN_Street_lighting_2007}.

@@

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication.\removed{We further propose that} Such a routing protocol can be used\rev{, for example,} in a smart street lighting system \removed{based on IPv6-enabled} wireless \removed{sensor}network for supporting diverse applications in an urban IoT. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

\section{Requirements and benefits of a smart street lighting system}
\label{secRequirements}
A smart street lighting system based on a network that covers entire cities is a widespread concept in the literature \citep{Our_IAS_2009, WSN_Street_Lighting_2010_1, Cellular_Street_Lighting_2010_1, Pantoni_2011, daFonseca_2015}. However, the requirements for such network are not always clearly defined. Despite that, through a simple analysis it is clear that most of the messages in this system will be towards the sink node, generating a converging traffic. As the majority of the WSN protocols are designed for such traffic (e.g., the RPL protocol), we are led to think that there are few possible contributions that researchers can make to improve the network data flow. Therefore, it is important to express a more complete set of requirements that such a system demands for a communication network, stating the main goals of a smart street lighting system. Based on previous works cited above, we can assert that such goals are usually monitoring tasks, such as measuring current, voltage, energy consumption and state of the lamp/ballast. All this kind of message are sporadic or may be distributed in time to not overload the network, which has limited bandwidth due to the IEEE 802.15.4 standard. Thus, it is not possible to support the need for high performance P2P communication only by the requirements of a smart street lighting system.

@@

6LoWPAN is a protocol definition, standardized in Request for Comments (RFCs) 4944 and 6282 \citep{RFC4944,RFC6282}, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a MTU (minimum transmission unit) of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

@@

Even, there are problems related to the DAO operating modes. In non-storing mode, RPL requires the use of IPv6 source routing mechanism. Therefore, a Source Routing Header (SRH) for use strictly between RPL routers in the same RPL routing domain was proposed in RFC 6554. However, RPL source routing implements a policy where each IPv6 hop between the source and destination of the source route is specified within the SRH. Such approach leads to a large overhead and increases the consumption of the already scarce bandwidth. Further, as the IEEE 802.15.4 standard has a limited maximum transmission unit (MTU), that may also imply in increased fragmentation. Therefore, successful delivery of an IP packet depends on the successful delivery of all possible fragments, which are not desirable once the LLNs are generally characterized by high packet loss rates.

@@

In storing mode, each router has to be able to store all the routes downwards its position in the formed DODAGs. This implies that, for routers near to the root node, the required storage is only bounded by the number of paths to all other routers in the network \citep{RPL_observations}. As RPL targets networks consisting of thousands of memory-constrained devices, the required storing capacity on these devices may not be compatible with the basic precepts of the WSANs.

Due to these not negligible drawbacks, it is wise to develop some work around able to improve the RPL protocol deficiencies, mainly that related with P2P messages. To this end, a reactive distance vector routing protocol called LOADng (Lightweight Ad hoc On-Demand - Next Generation (LOADng)) has been proposed \citep{LOADng_2012}. However, in recently published papers (\citep{Tripathi2014, Vucinic_2013}, LOADng was compared to RPL  and, according to the authors, LOADng may even have worse performance than RPL according to several metrics, including average delay, control overhead and maximum memory requirements. Therefore, in this paper, we follow another possible approach for routing. Such approach is the use of geographic routing techniques, which are routing algorithms known by its low overhead and high scalability features \citep{Karp_GPSR_2000, survey_2007, GOAFR_2008}.

@@

Many researchers argue that a dense deployment of wireless nodes can reduce the likelihood of the void occurrence in the network \citep{WSN_routing, survey_2007}. Indeed, in scenarios without obstacles that can be true. However, in outdoor large scale wireless networks it is unlikely to have such scenarios. Void handling techniques is an important issue to be addressed in such networks.

@@

\bibliographystyle{model2-names}
\bibliography{refs}

>>>>

The use of LED \rev{(light emitting diode)} lamps in street lighting started to be reported in the last few years, and it is being considered a very good solution to improve the energy efficiency of the system and also to reduce the maintenance cost, because of their long lifetime, high luminous efficiency and high color rendering index (CRI) \citep{CRI_2010}. Also, the addition of communication capability to the SLS decreases the energy consumption and the maintenance cost, because it allows for monitoring the status information of each lamp (state, current, voltage and power measurements, etc.) and to have better control of the system (e.g.: lamp dimming). Then, a smart SLS, based on LED lamps and a communication network, is an interesting application in a smart city context, not only by its clearly advantages but also because it can be one of the foundations for a urban ``IoT'' due to its pervasiveness across the city \citep{Smart_City_Smart_Light_2013}. Therefore, it can also be an enabler for the aforementioned applications in the smart city.

However, a street lighting system features a large number of independent devices, with geographic distribution depending on the city streets. Then, adding communication capabilities to these devices requires a complex network topology, as well as interoperability, scalability, security, robustness, low cost, ease to use and maintenance. A few contributions in the literature suggest some technologies that can be used to control a street lighting system. Some use the power lines for data transmission (PLC) \citep{PLC_Street_Lighting_2006}, while others use wireless communication, such as cellular networks \citep{Cellular_Street_Lighting_2010_1}, wireless sensor networks \rev{(WSN)} \citep{WSN_Street_Lighting_2010_1} and both combined \citep{WSN_Street_lighting_2007}.

@@

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication.\removed{We further propose that} Such a routing protocol can be used\rev{, for example,} in a smart street lighting system \removed{based on IPv6-enabled} wireless \removed{sensor}network for supporting diverse applications in an urban IoT\footnote{\rev{Throughout the text we have used the smart street lighting as an example of application where the proposed protocol could be used, however it was not specifically designed for a street lighting system, and thus might not fulfill all the requirements for such use.}}. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

\section{Requirements and benefits of a smart street lighting system}
\label{secRequirements}
A smart street lighting system based on a network that covers entire cities is a widespread concept in the literature \citep{Cellular_Street_Lighting_2010_1, WSN_Street_Lighting_2010_1, Denardin2009, Pantoni_2011, daFonseca_2015}. However, the requirements for such network are not always clearly defined. Despite that, through a simple analysis it is clear that most of the messages in this system will be towards the sink node, generating a converging traffic. As the majority of the WSN protocols are designed for such traffic (e.g., the RPL protocol), we are led to think that there are few possible contributions that researchers can make to improve the network data flow. Therefore, it is important to express a more complete set of requirements that such a system demands for a communication network, stating the main goals of a smart street lighting system. Based on previous works cited above, we can assert that such goals are usually monitoring tasks, such as measuring current, voltage, energy consumption and state of the lamp/ballast. All this kind of message are sporadic or may be distributed in time to not overload the network, which has limited bandwidth due to the IEEE 802.15.4 standard. Thus, it is not possible to support the need for high performance P2P communication only by the requirements of a smart street lighting system.

@@

6LoWPAN is a protocol definition, standardized in Request for Comments (RFCs) 4944 and 6282 \citep{RFC4944,RFC6282}, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a \removed{MTU} minimum transmission unit of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

@@

Even, there are problems related to the DAO operating modes. In non-storing mode, RPL requires the use of IPv6 source routing mechanism. Therefore, a Source Routing Header (SRH) for use strictly between RPL routers in the same RPL routing domain was proposed in RFC 6554. However, RPL source routing implements a policy where each IPv6 hop between the source and destination of the source route is specified within the SRH. Such approach leads to a large overhead and increases the consumption of the already scarce bandwidth. Further, as the IEEE 802.15.4 standard has a limited maximum transmission unit\removed{(MTU)}, that may also imply in increased fragmentation. Therefore, successful delivery of an IP packet depends on the successful delivery of all possible fragments, which are not desirable once the LLNs are generally characterized by high packet loss rates.

@@

In storing mode, each router has to be able to store all the routes downwards its position in the formed DODAGs. This implies that, for routers near to the root node, the required storage is only bounded by the number of paths to all other routers in the network \citep{RPL_observations}. As RPL targets networks consisting of thousands of memory-constrained devices, the required storing capacity on these devices may not be compatible with the basic precepts of the WSNs.

Due to these not negligible drawbacks, it is wise to develop some work around able to improve the RPL protocol deficiencies, mainly that related with P2P messages. To this end, a reactive distance vector routing protocol called LOADng (Lightweight Ad hoc On-Demand - Next Generation (LOADng)) has been proposed \citep{LOADng_2012}. However, in recently published papers (\citep{Tripathi2014, Vucinic_2013}, LOADng was compared to RPL  and, according to the authors, LOADng may even have worse performance than RPL according to several metrics, including average delay, control overhead and maximum memory requirements. Therefore, in this paper, we follow another possible approach for routing. Such approach is the use of geographic routing techniques, which are routing algorithms known by its low overhead and high scalability features \citep{Karp_GPSR_2000, GOAFR_2008, survey_2007}.

@@

Many researchers argue that a dense deployment of wireless nodes can reduce the likelihood of the void occurrence in the network \citep{survey_2007,WSN_routing}. Indeed, in scenarios without obstacles that can be true. However, in outdoor large scale wireless networks it is unlikely to have such scenarios. Void handling techniques is an important issue to be addressed in such networks.

@@

\bibliographystyle{elsarticle-num}
\bibliography{refs}
